---
title: "Reproducible Research Peer Assessment 1"
author: "Prashant Bajpai"
date: "Friday, July 18, 2014"
output: 
html_document:
toc: true        
theme: cerulean
---

## Loading and Preprocessing

Unzip the activity.zip file and load it into R. Remove rows having NA's

```{r}
setwd("D:/R projects//Learning/activity monitoring/")
mydata <- read.csv("activity.csv")
dat <- mydata[complete.cases(mydata),]
```

## Mean total number of steps taken per day


```{r, echo=TRUE}
library(knitr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(rCharts)

dt <- tbl_df(dat)
res <- dt %>% 
        group_by(date) %>%
        summarise(total.steps = sum(steps), mean = mean(steps), median = median(steps))

plot1<- ggplot(res, aes(x= res$date, y = total.steps)) + geom_histogram(fill = "#FF6666", color = "black", stat = "identity") + xlab("date") + scale_x_discrete(breaks = res$date[seq(1, length(res$date), by = 3)]) +  labs(title = "Mean steps per day after removing NA's") + theme(plot.title = element_text(lineheight = 0.8, face = "bold"))
```
```{r, fig.width=15}
plot1
```

Here is the table of containing mean and median of steps taken per day
```{r, results='asis'}
knitr::kable(res)
```

## Average daily activity pattern


```{r, fig.width=15}
dt$interval <- as.factor(dt$interval)
res2 <- dt %>%
        group_by(interval) %>%        
        summarise(average.steps = mean(steps))        

plot2 <- ggplot(res2, aes(x = interval, y = average.steps)) + geom_line(aes(group =1)) + scale_x_discrete(breaks = res2$interval[seq(1, length(res2$interval), by = 10)]) + labs(title = "Daily activity pattern after removing NA's") + theme(plot.title = element_text(lineheight = 0.8, face = "bold"))
plot2

maxsteps <- res2[which(res2$average.steps == max(res2$average.steps)),]$interval
```

The maximum number of steps averaged across all days were taken on **`r maxsteps`** interval

## Imputing missing values


```{r}
# Calculate missing values in mydata
missingvalues <- length(which(is.na(mydata$steps)))
```

The total number of rows with missing values are **`r missingvalues`**

Here I have replaced the missing values with average of total no. of steps for each interval  

```{r, fig.width=15}
compiled <- merge(mydata, res2, by = 'interval') 

for(i in 1:nrow(compiled)){
        if(is.na(compiled[i,2])){
        compiled[i,2] <- compiled[i,4]
        }
}

#calculate total number of steps taken daily along with mean and median
res3 <- compiled %>%
        group_by(date) %>%
        summarise(total.steps = sum(steps), mean = mean(steps), median = median(steps))


ggplot(res3, aes(x= date, y = total.steps)) + geom_histogram(fill = "#FF6666", color = "black", stat = "identity") + xlab("date") + scale_x_discrete(breaks = res$date[seq(1, length(res$date), by = 3)])



```

To see the effect of imputing NA's on lets first check the number of entries in tables before and after it. First dataset has **`r nrow(res)`** rows and after imputing we get **`r nrow(res3)`** rows. We lost information on some days by removing NA's. Apart from that I don't think it made a large difference.
Lets also check on the daily activity pattern with and without NA's removal

```{r, fig.width=15}
res4 <- compiled %>%
        group_by(interval) %>%        
        summarise(average.steps = mean(steps))    
plot3 <- ggplot(res4, aes(x = interval, y = average.steps)) + geom_line(aes(group =1)) + scale_x_discrete(breaks = res4$interval[seq(1, length(res4$interval), by = 10)]) + labs(title = "Daily activity pattern after imputing NA's") + theme(plot.title = element_text(lineheight = 0.8, face = "bold")) 
plot2
plot3
```

## Activity patterns between weekdays and weekends


Lets start with the dataframe in which NA's have been replaced by their subsequent means

```{r}
#Add a factor column of weekdays and weekend
compiled$date <- weekdays(as.Date(compiled$date))
myswitch <- function(x){
        switch(x, "Sunday" = "Weekend", "Monday" = "Weekday", "Tuesday" = "Weekday",
               "Wednesday" = "Weekday", "Thursday" = "Weekday", "Friday" = "Weekday",
               "Saturday" = "Weekend")
        }
for(i in 1:length(compiled$date)){
     compiled$date[i] <- myswitch(compiled$date[i])
}
compiled$date <- as.factor(compiled$date)
compiled$interval <- as.factor(compiled$interval)
compiled[,4] <- NULL

#splitting the data on th basis of date
mylist <- split(compiled, compiled$date)

#calculating mean for each interval
weekday <- mylist[[1]] %>%
        group_by(interval) %>%
        summarise(mean.steps = mean(steps)) %>%
        mutate(days = "Weekday")

weekend <- mylist[[2]] %>%
        group_by(interval) %>%
        summarise(mean.steps = mean(steps)) %>%
        mutate(days = "Weekend")

res5 <- rbind(weekday, weekend)
```
Here is the final figure separeted by weekday and weekend
```{r, fig.width=15}

ggplot(res5, aes(x = interval, y = mean.steps)) + 
        geom_line(aes(group =1), colour="blue") + 
        facet_grid(days~.) + 
        scale_x_discrete(breaks = res5$interval[seq(1, 288, by = 10)]) 
```
